censusprocessing

### 2021-11-29 code/analysis_code/analysis.Rmd
project pt 5

#load packages and data
```{r}
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) #for modeling
library(gridExtra) #for exporting tables

@@

# fit linear model to continuous outcome (particles_l) with log transformation
```{r}
#attempt to visualize data
#look at particles per L vs population
m3 <- ggplot(data, aes(x = population, y = particles_l)) + geom_point() +
  scale_y_continuous(trans='log10') + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. Population", x = "Population", y = "Log(Particles/Liter)")

@@

# particles per liter vs e.coli.cfu with log transformation
```{r}
m5 <- ggplot(data, aes(x = e.coli.cfu, y = particles_l)) + geom_point() + scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. E. coli CFU", x = "Log(CFU/mL)", y = "Log(Particles/Liter)")

@@

Trying code from above again with the complete dataset
# Modeling with complete cases only
```{r}
#estimating/training linear model
lm_fit4 <- lm_mod %>% fit(particles_l ~ ., data = complete)
#view summary of fit
tidy(lm_fit4)
#predict particles per liter based on population
lm_fit5 <- lm_mod %>% fit(particles_l ~ population, data = complete)
#view summary of fit
tidy(lm_fit5)

@@

Here, we have a VERY large RSS, which I think indicates a poor model fit for our data. Let's try with some variables removed. If an RSS of 0 is a perfect model, our RSS of 2 million is either a *very bad* model or I am making some sort of error/misinterpreting(?).

@@

#visualize linear model
t1 <- ggplot(data, aes(x = turbidity.ntu, y = particles_l)) + geom_point() + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. Turbidity", x = "Turbidity (ntu)", y = "Particles/Liter")
t1
#save figure
ggsave(filename = here("results","m_concvturbidity.png"), plot=t1)
```

@@

>>>>

#load packages and data
```{r}
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) #for modeling
library(gridExtra) #for exporting tables
library(gridExtra) #for savings images in grids

@@

# fit linear model to continuous outcome (particles_l) with log transformation
```{r}
#attempt to visualize data
#look at particles per L vs population
m3 <- ggplot(data, aes(x = population, y = particles_l)) + geom_point() +
  scale_y_continuous(trans='log10') + geom_smooth(method = lm) + labs(title = "MP Concentration vs. Population", x = "Population", y = "Log(Particles/Liter)")

@@

# particles per liter vs e.coli.cfu with log transformation
```{r}
m5 <- ggplot(data, aes(x = e.coli.cfu, y = particles_l)) + geom_point() + scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') + geom_smooth(method = lm) + labs(title = "MP Concentration vs. E. coli CFU", x = "Log(CFU/mL)", y = "Log(Particles/Liter)")

@@

Let's try the code from above again with the complete dataset

# Modeling with complete cases only
```{r}
#estimating/training linear model
lm_fit4 <- lm_mod %>% fit(particles_l ~ ., data = complete)
#view summary of fit
tidy(lm_fit4)
#predict particles per liter based on population
lm_fit5 <- lm_mod %>% fit(particles_l ~ population, data = complete)
#view summary of fit
tidy(lm_fit5)

@@

@@

#visualize linear model
t1 <- ggplot(data, aes(x = turbidity.ntu, y = particles_l)) + geom_point() + geom_smooth(method = lm) + labs(title = "MP Concentration vs. Turbidity", x = "Turbidity (ntu)", y = "Particles/Liter")
t1
#save figure
ggsave(filename = here("results","m_concvturbidity.png"), plot=t1)
```

@@

# linear model for distance from WRF
```{r}
#estimating/training linear model
wrf_fit1 <- lm_mod %>% fit(particles_l ~ dist, data = data)
#view summary of fit
tidy(wrf_fit1)

#visualize linear model
wrf1 <- ggplot(data, aes(x = dist, y = particles_l)) + geom_point() + geom_smooth(method = lm) + labs(title = "MP Concentration vs. WRF Distance", x = "WRF Distance (m)", y = "Particles/Liter")
wrf1
#save figure
ggsave(filename = here("results","m_concvwrfdist.png"), plot=wrf1)
```
# save grid image of results
```{r}
# arrange grid with population, e coli, WRF, and turbidity predictive models
bivar_grid <- grid.arrange(m3, m5, t1, wrf1, ncol = 2)

# save and export image
ggsave(bivar_grid, filename = here("results", "bivar_grid.png"))
```

### 2021-11-29 code/analysis_code/exploration.Rmd
project pt 5

# load packages and data
```{r}
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse)
library(ggplot2)
library(GGally) #for producing correlation matrices
library(gridExtra) #

>>>>

# load packages and data
```{r}
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse)
library(ggplot2)
library(GGally) #for producing correlation matrices
library(gridExtra) #
library(gtsummary)

### 2021-11-29 code/analysis_code/land_model.Rmd
project pt 5

## LASSO model
```{r}
#based on tidymodels tutorial: case study
#cross validation
set.seed(123)
cell_folds <- vfold_cv(train_data)

#model specification
lasso <- linear_reg(penalty = tune()) %>% set_engine("glmnet") %>% set_mode("regression")

>>>>

## LASSO model
```{r}

#cross validation
set.seed(123)
cell_folds <- vfold_cv(train_data)

#model specification
#lasso <- linear_reg(penalty = tune()) %>% set_engine("glmnet") %>% set_mode("regression")
lasso <- linear_reg() %>%
  set_mode("regression") %>%           
  set_engine("glmnet") %>%
  set_args(penalty = tune(), mixture = 1)

### 2021-11-29 code/analysis_code/modeling.Rmd
project pt 5

# load packages and data
```{r}
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) #for modeling
library(rpart)
library(glmnet)
library(ranger)
library(rpart.plot)  # for visualizing a decision tree
library(vip)         # for variable importance plots
library(gridExtra)

@@

#model specification
lasso <- linear_reg(penalty = tune()) %>% set_engine("glmnet") %>% set_mode("regression")

@@

#model predictions from tuned model vs actual outcomes
lasso_pred_plot <- ggplot(lasso_resid, aes(x = particles_l, y = .pred)) + geom_point() + 
  labs(title = "LASSO: Pred vs Outcomes", x = "Particles/L Outcome", y = "Particles/L Prediction")
lasso_pred_plot

#plot residuals vs predictions
lasso_resid_plot <- ggplot(lasso_resid, aes(y = .resid, x = .pred)) + geom_point() + 
  labs(title = "LASSO: Pred vs Residuals", x = "Particles/L Prediction", y = "Residuals")
lasso_resid_plot #view plot

@@

#model predictions from tuned model vs actual outcomes
tree_pred_plot <- ggplot(tree_resid, aes(x = particles_l, y = .pred)) + geom_point() + 
  labs(title = "Decision Tree: Pred vs Outcomes", x = "Particles/L Outcome", y = "Particles/L Prediction")
tree_pred_plot

#plot residuals vs predictions
tree_resid_plot <- ggplot(tree_resid, aes(y = .resid, x = .pred)) + geom_point() + 
  labs(title = "Decision Tree: Pred vs Residuals", x = "Particles/L Prediction", y = "Residuals")
tree_resid_plot #view plot

@@

The best decision tree has an RMSE of 102.2, which is better than the null model RMSE of 109. However, the decision tree did not perform as well well as the Lasso regularized model, which had an RMSE of about 100.

@@

# set seed for reproducible analysis (instead of random subset each time)
set.seed(123)
#subset 3/4 of data as training set
data_split <- initial_split(complete, 
                            prop = 7/10, 
                            strata = particles_l) #stratify by MP concentration for balanced outcome

#save sets as data frames
train_data <- training(data_split)
test_data <- testing(data_split)

#reset recipe with adjusted, complete data
MP_rec <- recipe(particles_l ~ ., data = train_data) %>% step_dummy(all_nominal())
#reset cell_folds
set.seed(123)
cell_folds <- vfold_cv(train_data)
```

@@

#fit model with finalized WF
rf_fit <- rf_final_wf %>% fit(train_data)
```
Here, we get an error message that all models failed, note: missing value in visual_score. Let's go back and use only complete cases for the random forest.

@@

### Random forest plots
```{r}
#diagnostics
autoplot(r_forest_res)
#calculate residuals
rf_resid <- rf_fit %>%
  augment(train_data) %>% #this will add predictions to our df
  select(.pred, particles_l) %>%
  mutate(.resid = particles_l - .pred) #manually calculate residuals

#model predictions from tuned model vs actual outcomes
rf_pred_plot <- ggplot(rf_resid, aes(x = particles_l, y = .pred)) + geom_point() + 
  labs(title = "Random Forest: Pred vs Outcomes", x = "Particles/L Outcome", y = "Particles/L Prediction")
rf_pred_plot

#plot residuals vs predictions
rf_resid_plot <- ggplot(rf_resid, aes(y = .resid, x = .pred)) + geom_point() + 
  labs(title = "Random Forest: Pred vs Residuals", x = "Particles/L Prediction", y = "Residuals")
rf_resid_plot #view plot

@@

```{r}
model_grid <- grid.arrange(lasso_pred_plot, lasso_resid_plot, rf_pred_plot, rf_resid_plot, tree_pred_plot, tree_resid_plot, ncol = 3)

@@

I'll need to go back and edit this image later so that the titles aren't cut off and the axis scales match up properly so the plots look right. I might also want to add a 45 degree diagonal line to the outcomes vs predictions plot, and some other features to improve appearance.

>>>>

# load packages and data
```{r}
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) #for modeling
library(rpart)
library(glmnet)
library(ranger)
library(rpart.plot)  # for visualizing a decision tree
library(vip)         # for variable importance plots
library(gridExtra) # for savings images in grids

@@

#model specification
#lasso <- linear_reg(penalty = tune()) %>% set_engine("glmnet") %>% set_mode("regression")
lasso <- linear_reg() %>%
  set_mode("regression") %>%           
  set_engine("glmnet") %>%
  set_args(penalty = tune(), mixture = 1)

@@

#model predictions from tuned model vs actual outcomes
lasso_pred_plot <- ggplot(lasso_resid, aes(x = particles_l, y = .pred)) + geom_point() + 
  labs(title = "LASSO", x = "Outcome (Particles/L)", y = "Prediction (Particles/L)")+ ylim(0, 1000) + xlim(0,1000)
lasso_pred_plot
ggsave(lasso_pred_plot, filename = here("results", "lasso_pred.png"))

#plot residuals vs predictions
lasso_resid_plot <- ggplot(lasso_resid, aes(y = .resid, x = .pred)) + geom_point() + 
  labs(title = "LASSO", x = "Prediction (Particles/L)", y = "Residuals")
lasso_resid_plot #view plot

@@

#model predictions from tuned model vs actual outcomes
tree_pred_plot <- ggplot(tree_resid, aes(x = particles_l, y = .pred)) + geom_point() + 
  labs(title = "Decision Tree", x = "Outcome (Particles/L)", y = "Prediction (Particles/L)")+ ylim(0, 1000) + xlim(0,1000)
tree_pred_plot

#plot residuals vs predictions
tree_resid_plot <- ggplot(tree_resid, aes(y = .resid, x = .pred)) + geom_point() + 
  labs(title = "Decision Tree", x = "Prediction (Particles/L)", y = "Residuals")
tree_resid_plot #view plot

@@

The best decision tree has an RMSE of 102.2, which is better than the null model RMSE of 109. However, the decision tree did not perform as well well as the Lasso regularized model, which had an RMSE of about 100. Based on the outcome/prediction plot, it appears that the model is making one of two predictions for any given actual outcome. This indicates that this is not a good model, as we don't see the expected 45 degree diagonal line where predictions correlate closely with outcomes.

@@

# set seed for reproducible analysis (instead of random subset each time)
set.seed(123)
#subset 3/4 of data as training set
complete_data_split <- initial_split(complete, 
                            prop = 7/10, 
                            strata = particles_l) #stratify by MP concentration for balanced outcome

#save sets as data frames
complete_train_data <- training(complete_data_split)
complete_test_data <- testing(complete_data_split)

#reset recipe with adjusted, complete data
MP_rec <- recipe(particles_l ~ ., data = complete_train_data) %>% step_dummy(all_nominal())
#reset cell_folds
set.seed(123)
cell_folds <- vfold_cv(complete_train_data)
```

@@

#fit model with finalized WF
rf_fit <- rf_final_wf %>% fit(complete_train_data)
```
Here, we get an error message that all models failed, note: missing value in visual_score. Let's go back and use only complete cases for the random forest.

@@

### Random forest plots
```{r}
#diagnostics
autoplot(r_forest_res)
#calculate residuals
rf_resid <- rf_fit %>%
  augment(complete_train_data) %>% #this will add predictions to our df
  select(.pred, particles_l) %>%
  mutate(.resid = particles_l - .pred) #manually calculate residuals

#model predictions from tuned model vs actual outcomes
rf_pred_plot <- ggplot(rf_resid, aes(x = particles_l, y = .pred)) + geom_point() + 
  labs(title = "Random Forest", x = "Outcome (Particles/L)", y = "Prediction (Particles/L)")+ ylim(0, 1000) + xlim(0,1000)
rf_pred_plot

#plot residuals vs predictions
rf_resid_plot <- ggplot(rf_resid, aes(y = .resid, x = .pred)) + geom_point() + 
  labs(title = "Random Forest", x = "Prediction (Particles/L)", y = "Residuals")
rf_resid_plot #view plot

@@

# Model comparison plots
```{r}
model_grid <- grid.arrange(lasso_pred_plot, rf_pred_plot, tree_pred_plot, lasso_resid_plot, rf_resid_plot, tree_resid_plot, ncol = 3)

@@

# Model selection : LASSO
```{r}
#Based on model performance plots and RMSE, we've selected the LASSO model. Let's evaluate it with the test data.

#fit to test data
last_lasso_fit <- lasso_final_wf %>% last_fit(data_split)
last_lasso_fit %>% collect_metrics()
```

It appears that the lasso model receives an RMSE of ~300 when run with the test data. Compared to the RMSE of 100 with the train data, this is not great performance and indicates that the model overfits the train data.

# Variable importance
```{r}
library(parsnip)
var_importance <- last_lasso_fit %>% 
  purrr::pluck(".workflow", 1) %>%   
  workflows::extract_fit_parsnip() %>% 
  vip(num_features = 20)

ggsave(var_importance, filename = here("results", "var_importance.png"))
```

# Final lasso plots
```{r}
#diagnostics
autoplot(lasso_res)
#calculate residuals
last_lasso_resid <- last_lasso_fit %>%
  augment() %>% #this will add predictions to our df
  select(.pred, particles_l) %>%
  mutate(.resid = particles_l - .pred) #manually calculate residuals

#model predictions from tuned model vs actual outcomes
last_lasso_pred_plot <- ggplot(last_lasso_resid, aes(x = particles_l, y = .pred)) + geom_point() + 
  labs(title = "LASSO: Pred vs Outcomes", x = "Test Outcome (Particles/L)", y = "Test Prediction (Particles/L)")+ ylim(0, 250) + xlim(0,250)
last_lasso_pred_plot
ggsave(last_lasso_pred_plot, filename = here("results", "last_lasso_pred.png"))

#plot residuals vs predictions
last_lasso_resid_plot <- ggplot(last_lasso_resid, aes(y = .resid, x = .pred)) + geom_point() + 
  labs(title = "LASSO: Test Pred vs Residuals", x = "Test Prediction (Particles/L)", y = "Residuals")
last_lasso_resid_plot #view plot

#compare to null model
metrics_null_train #view null RMSE for train data
last_lasso_fit %>% collect_metrics() #view RMSE for best lasso model

### 2021-11-29 code/analysis_code/subset_modeling.Rmd
code files
project pt 5

In this portion of the analysis, we will perform modeling on a smaller subset of data in attempt to improve model performance.

@@

#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","processeddata.rds")

@@

## LASSO model
```{r}
#based on tidymodels tutorial: case study
#cross validation
set.seed(123)
cell_folds <- vfold_cv(train_data)

#model specification
lasso <- linear_reg(penalty = tune()) %>% set_engine("glmnet") %>% set_mode("regression")

#set workflow
lasso_wf <- workflow() %>% add_model(lasso) %>% add_recipe(MP_rec)

@@

```{r}
#check for Poisson distribution
#https://stackoverflow.com/questions/59809960/how-do-i-know-if-my-data-fit-a-poisson-distribution-using-r
dispersion_test <- function(x) 
{
  res <- 1-2 * abs((1 - pchisq((sum((x - mean(x))^2)/mean(x)), length(x) - 1))-0.5)

  cat("Dispersion test of count data:\n",
      length(x), " data points.\n",
      "Mean: ",mean(x),"\n",
      "Variance: ",var(x),"\n",
      "Probability of being drawn from Poisson distribution: ", 
      round(res, 3),"\n", sep = "")

  invisible(res)
}

set.seed(1)
x <- data$particles_l

dispersion_test(x)
```

@@

>>>>

In this portion of the analysis, we will perform modeling on a smaller subset of data in attempt to improve model performance. The original aim of this document was to perform feature selection/subsetting for model improvement. However, the implementation of the LASSO method eliminated the need for this by essentially automating subset selection. The contents of this document are needed for the final analysis/manuscript product, but are kept in the project repository for understanding of the whole analysis process.

@@

#path to data

@@

## LASSO model
```{r}

#cross validation
set.seed(123)
cell_folds <- vfold_cv(train_data)

#model specification
#lasso <- linear_reg(penalty = tune()) %>% set_engine("glmnet") %>% set_mode("regression")
lasso <- linear_reg() %>%
  set_mode("regression") %>%           
  set_engine("glmnet") %>%
  set_args(penalty = tune(), mixture = 1)
#set workflow
lasso_wf <- workflow() %>% add_model(lasso) %>% add_recipe(MP_rec)

@@

# Model selection : LASSO
```{r}
#We've selected the LASSO model. Let's evaluate it with the test data.

#fit to test data
last_lasso_fit <- lasso_final_wf %>% last_fit(data_split)
last_lasso_fit %>% collect_metrics()
```

It appears that the lasso model receives an RMSE of ~197 when run with the test data. Compared to the RMSE of 113 with the train data, this is not great performance and indicates that the model overfits the train data.

# Variable importance
```{r}
library(parsnip)
last_lasso_fit %>% 
  purrr::pluck(".workflow", 1) %>%   
  workflows::extract_fit_parsnip() %>% 
  vip(num_features = 20)

@@

This plot shows us the variable importance rankings for the LASSO model, and we see that the watershed identity of each each sample site has importance for predicting microplastics concentration, while the hypothesized predictors (population, distance from WWTP, turbidity, e.coli.cfu, visual score) do not have importance in this model.

### 2021-11-29 code/processing_code/censusprocessing.Rmd
project pt 5

This function produces the zip codes I am looking for, but there are observations with missing variables that didn't return addresses and zip codes. I'll have to work on troubleshooting this. I may be able to manually input the zip codes, but that is not ideal.

# renaming variable differentiate use in processing
```{r}
popdata <- combodata
#save data as RDS
saveRDS(popdata, file = here("data", "processed_data", "popdata.rds"))

>>>>

# renaming variable to differentiate use in processing
```{r}
popdata <- combodata
#save data as RDS
saveRDS(popdata, file = here("data", "processed_data", "popdata.rds"))
